---
title: "Bayesian differential cell type proportion analysis: Toy dataset example"
author: Tanya Karagiannis, Stefano Monti, Paola Sebastiani
output:
  html_document:
    theme: paper
    code_folding: show
    toc: true
    toc_float: true
    toc_collapse: false
---

# Introduction

Bayesian differential cell type proportion analysis: Bayesian multinomial logistic regression model initialization and example application to a generated toy dataset.

Install the model_multinom.R function by cloning the github repository and load the function in R.

```{r, include = FALSE}
#Set work directory and load libraries
work.dir <- "/restricted/projectnb/uh2-sebas/analysis/scCentenarians/Composition_Multinomial/RJAGS_Composition_Analysis/"

```

```{r, messages=FALSE, warnings=FALSE}
source('./model_multinom.R')
```

# Usage

In this example, we generate a toy dataset and apply the Bayesian multinomial logistic regression model to the distributions of 10 randomly generated samples (labelled 5 younger and 5 older) with max 10000 counts with 3 cell types with set proportions that add up to 1.

# Load libraries

```{r, message = FALSE}
library(tidyverse)
library(rjags)
library(coda)
library(hablar)
library(patchwork)
```


## Generate toy dataset
```{r}
set.seed(562)
# 5 younger samples with set probabilities (0.1, 0.2, 0.7) of 3 categories (i.e. cell types)
sim.data.Y <- t(rmultinom(5, size = 10000, prob = c(0.1,0.2,0.7)))
# 5 older samples with set probabilities (0.2, 0.3, 0.5) of 3 categories (i.e. cell types)
sim.data.O <- t(rmultinom(5, size = 10000, prob = c(0.2,0.3,0.5)))
#bind sample count together
sim.data <- rbind(sim.data.Y, sim.data.O)

#create an age variable with 3 younger samples and 3 older samples
age.group <- c(rep(1,5), rep(2,5))
#pull simulated data and age variable in jags data object
jags.data <- list(y = sim.data, #simulated counts
                  age.group = factor(age.group), #age group label (younger = 1, older = 2)
                  N.sample = nrow(sim.data), #number of samples
                  N.ct = ncol(sim.data), #number of cell types
                  N.total = rep(10000,10), #total number of counts per sample
                  N.age = 2, #number of age groups
                  sex = sample(1:2, 10, replace = TRUE) #sex label 
)
jags.data

```

## Run Bayesian multinomial logistic regression model

```{r}
#Run model with 500 burn-in iterations and 1000 total iterations
jags <- jags.model(textConnection(multinom.model),data=jags.data, n.adapt=500, inits = list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 10))
#monitor predicted probabilities p
test <- coda.samples(jags, c('p'), n.adapt = 500, n.iter=1000)

```

## Obtain predicted probabilities of cell types for each sample

```{r}
#Get estimates of p[i,j] for sample i and category j
coda.summary <- summary(test) 
predicted.prob <- coda.summary$statistics %>% as_tibble() %>% select(Mean)
predicted.prob <- matrix(as.vector(predicted.prob$Mean), nrow = 10, ncol = 3)
predicted.prob

#predicted probabilities for each sample should add up to 1
apply(predicted.prob,1,sum)
```

# Diagnostics for model convergence

## Density and trace plots

```{r, fig.dim = c(20,10)}
plot(test)
```

## Autocorrelation

```{r}
autocorr.plot(test)

```

## Geweke convergence diagnostic

```{r}
geweke.diag(test)

```

